\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-07 chapters/gaze_independence.tex
checking bttda

One of the goals of brain-computer interfacing is establishing a communication
channel that does not rely on speech or muscular activity which in turn can
provide solutions to individuals with \ac{sspi}.
In the strictest interpretation, this then means that an interface should not
rely on the control of eye muscles used to redirect the gaze or for blinking.
It is exactly this potential of \acp{bci} that makes them suitable as assistive
communication devices.

\subsection{Incidence of eye motor impairment}
\Ac{sspi} is often caused by damage to the central or peripheral nervous
system, either through congenital diseases (\ac{cp}, \ac{fa}, \ldots),
neurodegenerative (\ac{als}, \ac{ms}, \ldots) or acquired (stroke and
\ac{tbi}).
Many individuals in these groups unfortunately also suffer to
some extent from eye motor impairment, requiring \ac{bci}s adapted to their condition.
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor impairment, which can range from minor (nystagmus
\footnote{%
Involuntary, rhythmic, and repetitive eye movements recognizable by their
consistent directionality (horizontal, vertical, or rotational).
}, other eye tremors
\footnote{%
These can include square-wave jerks, saccadic intrusions, microtremors, or
microsaccades while resting or fixating the gaze.
}, gaze fixation fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia
\footnote{%
Weakness or limited paralysis of one or more of the muscles that control eye
movement, leading to restricted eye motion, but not complete paralysis.
Often, a specific movement direction (up-down, left-right) is preserved.
}, involuntary movements, impaired pursuit, \ldots), and even complete
ophtalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
This not only has an effect on vision and coordination but also
on their ability to operate a visual \ac{bci}~\cite{FriedOken2020}.

\begin{table}
  %\sffamily
  %\footnotesize
	%\centering
  \input{chapters/gaze_independence_incidence.tex}
  \caption[Incidence of eye motor impairment in selected \acs{bci} target
    patient populations.]{%
      Incidence of eye motor impairment in selected \ac{bci} target
      patient populations. \acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
      \acl{dmd}, \acs{sma}: \acl{sma}, \acs{cp}: \acl{cp}, \acs{lis}: \acl{lis}.
      $+$: frequent, $-$: infrequent.
    }
    \label{tab:incidence}
\end{table}

Among the most  severely affected patients are stroke
patients~\cite{Pollock2011,Rowe2019}, mostly
due to brainstem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
These patients might suffer from severe or complete eye motor impairment
from the onset of their condition.
Individuals recovering from stroke with severe eye motor impairment are
Furthermore, those with
Case study reports~\cite{Patterson1986,Graber2016} show that even in individuals
with \ac{lis} due to stroke, the group with a complete lack of eye motor
control is very small.

In \ac{als}, a progressive disease affecting motor neurons, oculomotor issues
are also fairly common.
Although eye movement is often cited as one of the longest preserved
capabilities in \ac{als}, show that minor issues are still
fairly common, especially for patients with bulbar onset
\ac{als}~\cite{Kang2018, Guo2022,Moss2012}.
Furthermore, \textcite{Hayashi1991} show that, as these patients progress past
the point of independent breathing, symptoms will eventually also involve
eye muscle paralysis.

Various forms of eye movement abnormalities also occur often in
\ac{ms}.
\ac{ms} is a neurodegenerative disease involving demyelination of nerves.
Eye motor abnormalities are especially well
studied~\cite{Mueri1985,Prasad2010,Castelnovo2016,Serra2018,Polet2020} and
often used as diagnostical tool.
These abnormalities can be minor or severe, seldom progressing to complete
paralysis.
However, \ac{ms} often comes with vision loss, further complicating interaction
with visual \acp{bci}.

\ac{fa} is a neurodegenerative disease involving loss of coordinated movement.
This almost always affects eye
movements~\cite{Fahey2008,Hocking2010,Furman1983}, with various forms of
involuntary movements and troubles pursuing or fixating targets.
They also gradually have more troubles speaking, but often retain some muscular
control.

@@

\newcommand\fnlis{\footnote{
Multiple definitions of \ac{lis} are encountered in
\ac{bci} and neurological literature.
Some definitions include only tetraplegic patients without eye movements used
for communications.
Others distinguish Complete Locked-in Syndrome (CLIS) with full body paralysis,
including no eye motor control at all, from a \ac{lis} state with some preserved eye
movements or minor motor output.
While some definitions only include stroke or \ac{tbi} patients with damage to
specific regions in the brain (midbrain, brainstem, or
cerebellum)~\cite{Smith2005},
it can also generally refer to the state of full body paralysis
or loss of muscle tone incurred in neurodegenerative diseases, combined with
the inability to speak, such as occurs in late stage ALS.
}}

\newcommand\fnwolpawcrit{\footnote{
\it``The first class consists of people who are truly totally locked-in (e.g.,
due to end-stage ALS or severe cerebral palsy), who have no remaining
useful neuromuscular control of any sort, including no eye movement.
\elide\ This class is very small. \elide\
The second class of potential \ac{bci} users comprises those who retain
a very limited capacity for neuromuscular control. This group includes
people who retain some useful eye movement or enough limb muscle
function to operate a single-switch system. Such control is often slow,
unreliable, or easily fatigued.
This group is much larger than the first.
}}

We believe it is in general not opportune to delve too deep in the underlying
ophthalmological and neurological mechanisms for each of the specific
conditions, or even the exact symptoms.
In clinical reality, every patient different symptoms, which lead to their own
unique set of preserved capabilities and visual skills.
From a solution-oriented \ac{bci} engineering point
of view, the etiology of the symptoms can be abstracted away.
Instead of categorizing patients we will refer to those that might benefit from
\ac{bci} assistive communication technology as with \ac{sspi}, in line with
the terminology used by~\textcite{FriedOken2020}.
If they have eye motor impairment that affects their use of visual \ac{bci} or
eye tracking solutions, we use the term \ac{sspgi} (see Figure~\ref{fig:gaze_independence/venn}).

\begin{figure}
  \input{figures/gaze_independence/lis_venn.tikz.tex}
  \caption[Venn diagram of potential users of \acs{bci} assistive technology.]{%
    Venn diagram of potential users of assistive technology.
    Patients with \acf{sspi} are the general target population.
    If eye gaze-based solutions are not properly suited for an individual,
    they have \acf{sspgi}.
    Individuals with \ac{lis} have no muscular control left, or is it is very limited,
    such as the ability communicate through a binary switching system.
    \Acp{bci} are especially relevant for the last two groups.
  }
  \label{fig:gaze_independence/venn}
\end{figure}

@@

In general, the group of locked-in individuals with complete loss of eye motor
function is very small.
Hence, it would be interesting to focus \ac{bci} development efforts on the
larger population of individuals with \ac{sspgi}  that currently slip through
the cracks of the assistive technology offer.
These are individuals whose severe eye motor impairment prevents them from
using eye tracking based solutions.
They currently use their remaining motor control to
communicate by indicating symbols on a letterboard with great effort,
or signal with upwards eye movements or blinks to confirm prompted letters.
They require a caregiver or relative to interpret their signals, and crave
the ability to communicate independently which is crucial to retaining an
acceptable quality of life.

@@

In this work, we will use the term \emph{gaze-independent} meaning
\emph{‘dealing explicitly with the fact that a user cannot control their
gaze.'}
In the context of a visual \ac{bci}, this means that the user's
visuospatial attention and their gaze do not necessarily coincide.
There are multiple approaches to implement gaze-independence in a \ac{bci}.
The next section highlights some noteworhy examples.

@@

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but often only do so when paired
with invasive recording.
Non-visual reactive paradigms that use e.g. auditory~\cite{Halder2010} and
somatosensory~\cite{Brouwer2010}
stimulation do not rely on gaze redirection exist but also result in lower information transfer
rates compared to visual paradigms.
While auditory and somatosensory stimulation on its own might yield poor
\ac{itr} \textcite{Yin2016} that it can provide added value in gaze-independent
settings when coupled with oddball stimulation in a hybrid paradigm.
Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94bits/min.
In general, their approach is interesting since they follow a philosophy that
relies on establishing as many communication pathways as possible.
Yet, auditory and somatosensory \acp{bci} suffer from increased mental effort in
operation and from user-dependent variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

\subsection{Gaze-independent visual stimulation}
\label{sec:gaze-independence/sota/visual}
This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
Lin2018} or covertly~\cite{,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
speller~\parencite{Acqualagna2011}.
The \ac{rsvp} paradigm in particular is a prime contender for a performant gaze-independent
\ac{bci}.
\textcite{Lin2018} reported an average online \ac{itr} of 20.259 bits/min using
a paradigm with subsequently flashing sets of three characters filling the field of view of the user.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp},
where targets or small sets of targets are shown consecutively.
The Hex-o-Spell, Geospell and GIBS Block Speller are spatially organized, while
variants of the \ac{rsvp} paradigm are temporally organized.
\textcite{Lin2018} follow the philosophy that spatially organized interface
have historically shown not to perform well in the presence of gaze impairment.
However, we argue that, spatially organized interfaces in general have a much higher
\ac{itr} in healthy controls.
For instance, a regular matrix-based speller can achive an \ac{itr} to up to
30bits/min.
With the right set of adaptations, spatial attention could potentionally also
bring added value to individuals with \ac{sspgi}.
While spatial interfaces are indeed more gaze-dependent than temporal
interfaces they also provide an extra channel that can be used to transfer
information, i.e.\ the location of the stimulus.

As an alternative to spatial attention, non-spatial visual attention (feature attention) can also be exploited, such as
attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
The \ac{rsvp} speller is already an example of this, relying on the user to
attend to attend the appearance of the intended character to type.
Visual stimulation paradigms relying on alternative types of attention
can modulate specific extra \ac{erp} components that either improve performance
because the embed extra information in the brain signal~\cite{Xu2022}
or because they are more sensitive to stimulation in the visual
periphery~\cite{Schaeff2012}.
Nevertheless, solely relying on alternative types of attenion can suffer too
from reduced information transfer rates~\cite{Chennu2013}.
Furthermore, the previously mentioned systems typically were only tested in
settings where the user was required to focus on a central fixation point while
selecting peripheral targets.
This entails that they still rely to some extent on
eye motor control, often necessitating central gaze fixation.

\subsection{Gaze-independent decoding}
\label{sec:gaze-independence/sota/decoding}
Thirdly, stimuli can be presented in a standard \ac{bci} paradigm, but visuospatial
attention can be decoded separately from gaze direction.
\textcite{Aloise2012} aimed to bridge the performance gap between covert and
overt VSA decoding performance.
They compared classical linear and non-linear \ac{erp} classifiers on a covert
attention oddball \ac{erp} paradigm dataset.
The results revealed no significant performance improvement in covert VSA
decoding for any of the investigated decoders.

More recent work has made advances in decoding lateralized covert
VSA harnessing the N2pc \ac{erp}
component~\cite{Thiery2016,Reichert2020b,Wang2022}
The amplitude of the N2pc component in the contralateral hemisphere, is directly modulated
by the location of covert \ac{vsa}.
Another approach is directly decoding covert shifts in \ac{vsa} from spectral
content~\cite{Tonin2013}.
These methods are promising, but to date often require a slower stimulation pace than or cannot display as
many targets as the classical \ac{erp} \ac{bci} paradigms.
As with the alternative modality approach, they are currently only helpful as
extra communication channels in a hybrid paradigm.
\textcite{Xu2016} reached an \ac{itr} of 23.56bits/min by detecting the N2pc
component while performing \ac{ssvep} stimulation.
\textcite{Egan2017} also showed an increase in \ac{ssvep} performance by
concurrently decoding the covert \ac{vsa} shift from spectral content, but
their study only used a few stimulation targets.
In general, we can state that gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
leveraging spatial attention remains underexplored.

@@

To summarize: for patients with eye motor impairment, gazing directly at a screen target may
be uncomfortable, impractical, or even impossible.
Hence, assistive devices that rely on eye tracking are often inefficient for
them.
Consequently, while \ac{bci}s hold great promise for these individuals, conventional
gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
control.
Therefore, the development of decoding strategies that account for covert
VSA becomes crucial in the pursuit of high-performance gaze-independent
\ac{bci}s.

The general goal of this PhD is to tackle the gaze-dependence in problem in the
visual oddball \ac{bci}.
Section~\ref{sec:gaze-independence/sota} showed that visual \ac{bci} paradigms form
a good candidate for a performant gaze-independent \ac{bci}.
While the central gaze fixation of most visual gaze-independent paradigms still
relies to some extent on eye motor control, we aim to circumvent this.
In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
visuospatial attention can be decoded separately from the position of gaze,
which do not necessarily need to coincide.

This leads us to adopt a specific approach: \emph{improve the performance of a
spatially organized visual oddball \ac{erp}-based brain
computer interface by using a suited decoding strategy}.
%Hence, our working hypothesis is as follows:
%\todo{rewrite, this is not what we tested in the patient paper after all}
%\begin{quote}
%	\textit{A spatially organized visual oddball \ac{erp} \ac{bci} with a suited decoder
%		can achieve a similar transfer rate in patient
%		populations with eye motor deficits than other gaze-independent alternatives
%    described in literature.}
%\end{quote}

To achieve our goal, we chose a suitable interface and innovate on decoder
development.
We collect data from healthy control subjects, which allow us to test our decoders.
Finally, the findings of these experiments will be verified in experiments with
individuals with \ac{sspgi}, including individuals with \ac{lis} evaluating
both performance and comfort.

This specific target is embedded within a broader goal of enabling
effective communication for eye-motor impaired \ac{lis} patients.
Ultimately, we wish to design a comfortable interface that allows them to maximally exploit
their residual gaze capabilities.
We believe this can be done with a non-invasive, spatially
organized visual \acp{bci} with high-\ac{itr}.
Finally, or efforts also aim to improve hand improving \ac{erp} decoding performance
in general, which will also contribute to their effectiveness in gaze-dependent
or gaze-independent settings.

@@

During this PhD, we explored different lines in decoding strategies,
trying to tackle several problems, be it general problems in \ac{erp} decoding
or those that arise specifically from gaze-independence.
Examples of this are the lack or decrease in amplitude of specific \ac{erp} components, and the increased
non-stationarity of the signal.
As mentioned, state-of-the-art decoders have poor performance in covert attention
settings.
The general goal is thus to design a machine learning classifier that represents
the \ac{erp} signal in a specific way, such that it becomes more robust to the problems
occurring in covert attention conditions.

\subsubsection{Regularized spatiotemporal beamforming}
Due to the decreased amplitude of the N1 and P3 components in covert attention
settings~\cite{Treder2010}, the \ac{snr} of
the \ac{erp} is lower than in overt attention settings.
Therefore, a straightforward way to reach satisfactory
gaze-independent decoding performance, might be by increasing overall \ac{erp}
decoding performance.

Therefore, we have improved upon an in-house developed, state-of-the art \ac{erp}
decoder, the spatiotemporal beamformer~\cite{Wittevrongel2016}, by reformulating
this classifier as a linear discrimination problem and
imposing regularizing constraints by structuring the noise covariance matrix
(STBF-struct).
Furthermore, these regularizing constraints impose temporal stationarity on
the background noise, yielding insights for our next
efforts to cope with the non-stationarity of the P3 signal component.
This approach has been published by~\textcite{VanDenKerchove2022} and will be
described in Chapter~\ref{sec:stbf-struct}.

@@

In the context of \ac{bci} decoding, extracting robust and discriminative
features from multidimensional neural data is critical.
Tensor decoding methods offers a powerful approach by preserving the multiway structure
of \ac{eeg} data while optimizing class separability.
Unlike traditional methods that flatten data, TDA operates directly on tensors,
making it particularly suited for \ac{bci} applications, where signals are structured
across multiple channels and time points.

A widely used method is \ac{hoda}~\cite{Phan2010}, which projects tensors onto subspaces
optimized for class separation, similar to \ac{lda}, but extended to multidimensional data.
This method has been applied to various \ac{bci} tasks, including \ac{erp} and
\ac{mi} decoding, enabling the extraction of class-relevant features while reducing
the data's dimensionality.

While \ac{hoda} is effective, it is limited by its reliance on a rigid core
structure, which may not fully capture the complexity of neural data in some
scenarios. To address this, \ac{bttda}
introduces a more flexible block-term structure.
\ac{bttda} iteratively extracts multiple blocks of
discriminative information, which can help in capturing more complex patterns
in neural data.
While \ac{bttda} is not specifically tailored to
gaze-independent \ac{bci} decoding, specific tensorizations of neural data can
offer advantages in coping with several obstacles arising from
gaze-independence.
The proposed \ac{bttda} method will be discussed in~\autoref{sec:bttda}

>>>>

One of the goals of brain-computer interfacing is establishing a communication
channel that does not rely on speech or muscular activity, which in turn can
provide solutions to individuals with \ac{sspi}.
In the strictest interpretation, this means that an interface should not
rely on the control of eye muscles used to redirect the gaze or for blinking.
It is exactly this potential of \acp{bci} that makes them suitable as assistive
communication devices.

\subsection{Incidence of eye motor impairment}
\Ac{sspi} is often caused by damage to the central or peripheral nervous
system, either through congenital diseases (\ac{cp}, \ac{fa}, \ldots),
neurodegenerative (\ac{als}, \ac{ms}, \ldots) or acquired (stroke and
\ac{tbi}).
Many individuals in these groups unfortunately also suffer
to some extent from eye motor impairment, requiring \acp{bci} adapted to their condition.
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor impairment, which can range from minor (nystagmus,
\footnote{Involuntary, rhythmic, and repetitive eye movements recognizable by their
consistent directionality (horizontal, vertical, or rotational).},
other eye tremors
\footnote{These can include square-wave jerks, saccadic intrusions, microtremors, or
microsaccades while resting or fixating the gaze.},
gaze fixation fatigue or discomfort, \ldots), to severe (partial ophthalmoplegia
\footnote{Weakness or limited paralysis of one or more of the muscles that control eye
movement, leading to restricted eye motion, but not complete paralysis.
Often, a specific movement direction (up-down, left-right) is preserved.},
involuntary movements, impaired pursuit, \ldots), and even complete
ophthalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
This not only affects vision and coordination but also
their ability to operate a visual \ac{bci}~\cite{FriedOken2020}.

\begin{table}
  \input{chapters/gaze_independence_incidence.tex}
  \caption[Incidence of eye motor impairment in selected \acs{bci} target
    patient populations.]{%
      Incidence of eye motor impairment in selected \ac{bci} target
      patient populations. \acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
      \acl{dmd}, \acs{sma}: \acl{sma}, \acs{cp}: \acl{cp}, \acs{lis}: \acl{lis}.
      $+$: frequent, $-$: infrequent.
    }
    \label{tab:incidence}
\end{table}

Among the most severely affected patients are stroke
patients~\cite{Pollock2011,Rowe2019}, mostly
due to brainstem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
These patients might suffer from severe or complete eye motor impairment
from the onset of their condition.
Case study reports~\cite{Patterson1986,Graber2016} show that even in individuals
with \ac{lis} due to stroke, the group with a complete lack of eye motor
control is very small.

In \ac{als}, a progressive disease affecting motor neurons, oculomotor issues
are also fairly common.
Although eye movement is often cited as one of the longest preserved
capabilities in \ac{als}, studies show that minor issues are still
fairly common, especially for patients with bulbar onset
\ac{als}~\cite{Kang2018, Guo2022,Moss2012}.
Furthermore, \textcite{Hayashi1991} show that as these patients progress past
the point of independent breathing, symptoms will eventually also involve
eye muscle paralysis.

Various forms of eye movement abnormalities also occur often in
\ac{ms}.
\Ac{ms} is a neurodegenerative disease involving demyelination of nerves.
Eye motor abnormalities are especially well
studied~\cite{Mueri1985,Prasad2010,Castelnovo2016,Serra2018,Polet2020} and
are often used as diagnostic tools.
These abnormalities can be minor or severe, seldom progressing to complete
paralysis.
However, \ac{ms} often comes with vision loss, further complicating interaction
with visual \acp{bci}.

\Ac{fa} is a neurodegenerative disease involving loss of coordinated movement.
This almost always affects eye
movements~\cite{Fahey2008,Hocking2010,Furman1983}, with various forms of
involuntary movements and trouble pursuing or fixating on targets.
They also gradually have more trouble speaking but often retain some muscular
control.

@@

\newcommand\fnlis{\footnote{Multiple definitions of \ac{lis} are encountered in
\ac{bci} and neurological literature.
Some definitions include only tetraplegic patients without eye movements used
for communications.
Others distinguish Complete Locked-in Syndrome (CLIS) with full body paralysis,
including no eye motor control at all, from a \ac{lis} state with some preserved eye
movements or minor motor output.
While some definitions only include stroke or \ac{tbi} patients with damage to
specific regions in the brain (midbrain, brainstem, or
cerebellum)~\cite{Smith2005}, it can also generally refer to the state of full body paralysis
or loss of muscle tone incurred in neurodegenerative diseases, combined with
the inability to speak, such as occurs in late-stage ALS.}}

\newcommand\fnwolpawcrit{\footnote{
``The first class consists of people who are truly totally locked-in (e.g.,
due to end-stage ALS or severe cerebral palsy), who have no remaining
useful neuromuscular control of any sort, including no eye movement.
\elide\ This class is very small. \elide\
The second class of potential \ac{bci} users comprises those who retain
a very limited capacity for neuromuscular control. This group includes
people who retain some useful eye movement or enough limb muscle
function to operate a single-switch system. Such control is often slow,
unreliable, or easily fatigued.
This group is much larger than the first.''~\cite{Wolpaw2008}}}

We believe it is generally not opportune to delve too deeply into the underlying
ophthalmological and neurological mechanisms for each of the specific
conditions, or even the exact symptoms.
In clinical reality, every patient has different symptoms, which lead to their own
unique set of preserved capabilities and visual skills.
From a solution-oriented \ac{bci} engineering point of view, the etiology of the symptoms can be abstracted away.
Instead of categorizing patients, we will refer to those who might benefit from
\ac{bci} assistive communication technology as having \ac{sspi}, in line with
the terminology used by~\textcite{FriedOken2020}.
If they have eye motor impairment that affects their use of visual \ac{bci} or
eye-tracking solutions, we use the term \ac{sspgi} (see Figure~\ref{fig:gaze_independence/venn}).

\begin{figure}
  \input{figures/gaze_independence/lis_venn.tikz.tex}
  \caption[Venn diagram of potential users of \acs{bci} assistive technology.]{%
    Venn diagram of potential users of assistive technology.
    Patients with \acf{sspi} are the general target population.
    If eye gaze-based solutions are not properly suited for an individual,
    they have \acf{sspgi}.
    Individuals with \ac{lis} have no muscular control left, or it is very limited,
    such as the ability to communicate through a binary switching system.
    \Acp{bci} are especially relevant for the last two groups.
  }
  \label{fig:gaze_independence/venn}
\end{figure}

@@

In general, the group of locked-in individuals with complete loss of eye motor
function is very small.
Hence, it would be interesting to focus \ac{bci} development efforts on the
larger population of individuals with \ac{sspgi} who currently slip through
the cracks of the assistive technology offerings.
These are individuals whose severe eye motor impairment prevents them from
using eye-tracking-based solutions.
They currently use their remaining motor control to
communicate by indicating symbols on a letterboard with great effort,
or signal with upward eye movements or blinks to confirm prompted letters.
They require a caregiver or relative to interpret their signals and crave
the ability to communicate independently, which is crucial to retaining an
acceptable quality of life.

@@

In this work, we will use the term \emph{gaze-independent} to mean
\emph{`dealing explicitly with the fact that a user cannot control their
gaze.'}
In the context of a visual \ac{bci}, this means that the user's
visuospatial attention and their gaze do not necessarily coincide.
There are multiple approaches to implement gaze-independence in a \ac{bci}.
The next section highlights some noteworthy examples.

@@

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \acp{bci}~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but they often do so only when paired
with invasive recording.
Non-visual reactive paradigms that use, for example, auditory~\cite{Halder2010} and
somatosensory~\cite{Brouwer2010}
stimulation do not rely on gaze redirection but also result in lower information transfer
rates compared to visual paradigms.
While auditory and somatosensory stimulation on their own might yield poor
\ac{itr}, \textcite{Yin2016} showed that it can provide added value in gaze-independent
settings when coupled with oddball stimulation in a hybrid paradigm.
Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94 bits/min.
In general, their approach is interesting as it follows a philosophy that
relies on establishing as many communication pathways as possible.
Yet, auditory and somatosensory \acp{bci} suffer from increased mental effort in
operation and from user-dependent variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

\subsection{Gaze-independent visual stimulation}
\label{sec:gaze-independence/sota/visual}

This leads us to the second approach: current visual stimulation paradigms can be optimized so that
the stimuli are always present in the field of view, either overtly~\cite{Acqualagna2013, Won2018, Lin2018} or covertly~\cite{Pires2011,Lees2018}.
Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
speller~\parencite{Acqualagna2011}.
The \ac{rsvp} paradigm, in particular, is a prime contender for a performant gaze-independent
\ac{bci}.
\textcite{Lin2018} reported an average online \ac{itr} of 20.259 bits/min using
a paradigm with subsequently flashing sets of three characters filling the user's field of view.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp}, where targets or small sets of targets are shown consecutively.
The Hex-o-Spell, GeoSpell, and GIBS Block Speller are spatially organized, while
variants of the \ac{rsvp} paradigm are temporally organized.
\textcite{Lin2018} follow the philosophy that spatially organized interfaces
have historically not performed well in the presence of gaze impairment.
However, we argue that spatially organized interfaces generally have a much higher
\ac{itr} in healthy controls.
For instance, a regular matrix-based speller can achieve an \ac{itr} of up to
30 bits/min.
With the right set of adaptations, spatial attention could potentially also
bring added value to individuals with \ac{sspgi}.
While spatial interfaces are indeed more gaze-dependent than temporal
interfaces, they also provide an extra channel that can be used to transfer
information, i.e., the location of the stimulus.

As an alternative to spatial attention, non-spatial visual attention (feature attention) can also be exploited, such as
attention to stimulus color, shape, or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
The \ac{rsvp} speller is already an example of this, relying on the user to
attend to the appearance of the intended character to type.
Visual stimulation paradigms relying on alternative types of attention
can modulate specific extra \ac{erp} components that either improve performance
because they embed extra information in the brain signal~\cite{Xu2022}
or because they are more sensitive to stimulation in the visual
periphery~\cite{Schaeff2012}.
Nevertheless, solely relying on alternative types of attention can also suffer from
reduced information transfer rates~\cite{Chennu2013}.
Furthermore, the previously mentioned systems were typically tested only in
settings where the user was required to focus on a central fixation point while
selecting peripheral targets.
This entails that they still rely to some extent on
eye motor control, often necessitating central gaze fixation.

\subsection{Gaze-independent decoding}
\label{sec:gaze-independence/sota/decoding}

Thirdly, stimuli can be presented in a standard \ac{bci} paradigm, but visuospatial
attention can be decoded separately from gaze direction.
\textcite{Aloise2012} aimed to bridge the performance gap between covert and
overt \ac{vsa} decoding performance.
They compared classical linear and non-linear \ac{erp} classifiers on a covert
attention oddball \ac{erp} paradigm dataset.
The results revealed no significant performance improvement in covert \ac{vsa}
decoding for any of the investigated decoders.

More recent work has made advances in decoding lateralized covert
\ac{vsa} by harnessing the N2pc \ac{erp}
component~\cite{Thiery2016,Reichert2020b,Wang2022}.
The amplitude of the N2pc component in the contralateral hemisphere is directly modulated
by the location of covert \ac{vsa}.
Another approach is directly decoding covert shifts in \ac{vsa} from spectral
content~\cite{Tonin2013}.
These methods are promising, but to date often require a slower stimulation pace or cannot display as
many targets as the classical \ac{erp} \ac{bci} paradigms.
As with the alternative modality approach, they are currently only helpful as
extra communication channels in a hybrid paradigm.
\textcite{Xu2016} reached an \ac{itr} of 23.56 bits/min by detecting the N2pc
component while performing \ac{ssvep} stimulation.
\textcite{Egan2017} also showed an increase in \ac{ssvep} performance by
concurrently decoding the covert \ac{vsa} shift from spectral content, but
their study only used a few stimulation targets.
In general, we can state that gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
leveraging spatial attention remains underexplored.

@@

To summarize, for patients with eye motor impairment, gazing directly at a screen target may
be uncomfortable, impractical, or even impossible.
Hence, assistive devices that rely on eye tracking are often inefficient for
them.
Consequently, while \acp{bci} hold great promise for these individuals, conventional
gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
control.
Therefore, the development of decoding strategies that account for covert
\ac{vsa} becomes crucial in the pursuit of high-performance gaze-independent
\acp{bci}.

The general goal of this PhD is to tackle the gaze-dependence problem in the
visual oddball \ac{bci}.
Section~\ref{sec:gaze-independence/sota} showed that visual \ac{bci} paradigms are a
good candidate for a performant gaze-independent \ac{bci}.
While the central gaze fixation of most visual gaze-independent paradigms still
relies to some extent on eye motor control, we aim to circumvent this.
In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
visuospatial attention can be decoded separately from gaze position, which do not necessarily need to coincide.

This leads us to adopt a specific approach: \emph{improve the performance of a
spatially organized visual oddball \ac{erp}-based brain
computer interface by using a suited decoding strategy}.

To achieve our goal, we chose a suitable interface and innovate on decoder
development.
We collect data from healthy control subjects to test our decoders.
Finally, the findings of these experiments will be verified in experiments with
individuals with \ac{sspgi}, including individuals with \ac{lis}, evaluating
both performance and comfort.

This specific target is embedded within a broader goal of enabling
effective communication for eye-motor impaired \ac{lis} patients.
Ultimately, we wish to design a comfortable interface that allows them to maximally exploit
their residual gaze capabilities.
We believe this can be done with a non-invasive, spatially
organized visual \acp{bci} with high \ac{itr}.
Finally, our efforts also aim to improve \ac{erp} decoding performance
in general, which will also contribute to their effectiveness in both gaze-dependent
and gaze-independent settings.

@@

During this PhD, we explored different lines of decoding strategies,
trying to tackle several challenges, be they general problems in \ac{erp} decoding
or those that arise specifically from gaze-independence.
Examples of these challenges include the lack or decrease in the amplitude of specific \ac{erp} components, and the increased
non-stationarity of the signal.
As mentioned, state-of-the-art decoders have poor performance in covert attention
settings.
The general goal is thus to design a machine learning classifier that represents
the \ac{erp} signal in a way that makes it more robust to the problems
occurring in covert attention conditions.

\subsubsection{Regularized spatiotemporal beamforming}

Due to the decreased amplitude of the N1 and P3 components in covert attention
settings~\cite{Treder2010}, the \ac{snr} of
the \ac{erp} is lower than in overt attention settings.
Therefore, a straightforward way to reach satisfactory
gaze-independent decoding performance may be to increase overall \ac{erp}
decoding performance.

To address this, we improved upon an in-house developed, state-of-the-art \ac{erp}
decoder, the spatiotemporal beamformer~\cite{Wittevrongel2016}, by reformulating
this classifier as a linear discrimination problem and
imposing regularizing constraints by structuring the noise covariance matrix
(STBF-struct).
Furthermore, these regularizing constraints impose temporal stationarity on
the background noise, yielding insights for our next
efforts to cope with the non-stationarity of the P3 signal component.
This approach has been published by~\textcite{VanDenKerchove2022} and will be
described in \autoref{sec:stbf-struct}.

@@

In the context of \ac{bci} decoding, extracting robust and discriminative
features from multidimensional neural data is critical.
Tensor decoding methods offer a powerful approach by preserving the multiway structure
of \ac{eeg} data while optimizing class separability.
Unlike traditional methods that flatten data, TDA operates directly on tensors,
making it particularly suited for \ac{bci} applications, where signals are structured
across multiple channels and time points.

A widely used method is \ac{hoda}~\cite{Phan2010}, which projects tensors onto subspaces
optimized for class separation, similar to \ac{lda} but extended to multidimensional data.
This method has been applied to various \ac{bci} tasks, including \ac{erp} and
\ac{mi} decoding, enabling the extraction of class-relevant features while reducing
the data's dimensionality.

While \ac{hoda} is effective, it is limited by its reliance on a rigid core
structure, which may not fully capture the complexity of neural data in some
scenarios. To address this, \ac{bttda}
introduces a more flexible block-term structure.
\ac{bttda} iteratively extracts multiple blocks of
discriminative information, which can help in capturing more complex patterns
in neural data.
While \ac{bttda} is not specifically tailored to
gaze-independent \ac{bci} decoding, specific tensorizations of neural data can
offer advantages in coping with several obstacles arising from
gaze-independence.
The proposed \ac{bttda} method will be discussed in \autoref{sec:bttda}.

\subsubsection{Classifier-based Latency Estimation with Woody iterations}
